
\chapter{Enhancing the Creation of Mapping Rules}
\label{chapter:creation}

The mappings used for constructing knowledge graphs can be created in different ways. As they are formatted as text documents, they can be directly written in any text editor. However, the verbosity of some languages and the need of following a specific syntax has foster the development of user-friendly tools and serializations to ease the mapping writing process. This chapter introduces two solutions for easing the construction of mappings for different potential users, (i) a spreadsheet-based approach to write the mapping rules, and (ii) a user-friendly mapping syntax updated with the latest needs in knowledge graph construction.

\input{5_creation/5-1_mapeathor}

\input{5_creation/5-2_yarrrml-star}

\section{Conclusions}

This chapter addresses the second objective of this thesis \textit{O2: To help knowledge engineers and domain experts to build mappings documents providing the means for a user-friendly experience.}
%In this chapter, we extend the tooling assistance of mapping creation for users. 
The first contribution within this objective is \textit{C4: Design of a spreadsheet-based approach for writing mapping rules.} This contribution achieves a two-fold objective: (i) To ease the writing of mapping rules in a familiar environment (spreadsheets), while erasing the need of learning a mapping language syntax; and (ii) to increase the interoperability among languages, as the spreadsheet representation can be translated into more than one language. We test the developed approach in a user study with 30 volunteer practitioners, comparing it to a graphical approach (RMLEditor) and the baseline (writing mappings directly in the RML mapping language). The results suggest that participants manage to write mappings with higher quality and more complete than with the other approaches. Thus, this suggest that the spreadsheet-based approach is suitable for enhancing the mapping writing process for users. 

The second contribution tackles \textit{C5: Update of the a user-friendly syntax with extended features}. 
We present the latest updates to a user-friendly syntax for the RML mapping language, YARRRML, one of the most popular and adopted for writing RML mappings. 
We qualitatively validate the updates proposed by comparing them to other user-friendly syntaxes to ensure maximum expressiveness coverage. 

Finally present the implementations for both presented approaches, Mapeathor for generating [R2]RML and YARRRML mappings from spreadsheets, and Yatter for translating the updated YARRRML syntax into [R2]RML. The objective of the contributions presented in this chapter is to provide further support for user to ease the writing of mappings, aiming at reducing the barrier for adopting mappings to build knowledge graphs. 

%\ana{añadir otra contribución aquí con la interoperabilidad de las herramientas?}