
%\chapter{A Unified Model for Representing Declarative Mapping Languages}
\chapter{Understanding, Representing and Extending the Expressiveness of Declarative Mapping Languages}
\label{chapter:mappings}

Declarative mappings are a key element for the knowledge graph construction process to enhance maintainability, understandability and reproducibility. This chapter first presents an extensive analysis of current mapping languages in the form of a comparison framework (\cref{sec:chp4_framework}). Based on this comparison, a set of requirements are extracted and used for building an ontology that aims at representing the expressiveness of current mapping languages (\cref{sec:chp4_cm_ontology}). Finally, we present the extension of a widely used mapping language to construct RDF-star graphs (\cref{sec:chp4_rml_star}).

%\ana{al igual esta intro se puede extender más, explicando un poco más cada parte y situandola en su contexto, que no sea una enumeración de las subsecciones.}

\input{4_mappings/4-1_framework}

\input{4_mappings/4-2_cm_ontology}

\input{4_mappings/4-3_rml_star}

\section{Conclusions}

This chapter addresses the fist objective of this thesis \textit{O1:To analyze the needs for declarative knowledge graph construction from heterogeneous data sources, in order to facilitate the support of advances in existing mapping languages to address the most relevant ones.} 
To reach this objective, we first performed an in-depth analysis of the features of the current mapping languages: from the description of input data sources, to the diverse features for triple generation. This comprises the first contribution of \textit{O1, C1: Design of a comparison framework analysing the state of the art mapping languages.}
This analysis and the challenges proposed by the Knowledge Graph Construction W3C Community Group are gathered as requirements for a mapping language to be able to address the complexity of use cases for KG construction. 

Then, we presented the Conceptual Mapping, an ontology that, based on the abovementioned requirements, gathers the expressiveness of current mapping languages. This ontology was developed following the LOT methodology, it is evaluated accordingly and made available under a W3ID. 
This comprises the second contribution ob objective O1, \textit{C2: Identify, define and implement the requirements for knowledge graph construction from heterogeneous data sources.}

Finally, we show that mapping languages evolve as KG technologies advance, in this case with the adoption of RDF-star. To that purpose, we extend one of the most used mapping languages, RML, to create RDF-star graphs, RML-star. Accordingly, we add an extension to the Conceptual Mapping to include this new feature. This comprises the third contribution of objective O1, \textit{C3 Development of new features for the RML mapping language.}

Our ambition with these contributions is to enhance the understanding of the capabilities of mapping languages and what they need to implement in order to cover the necessities for constructing knowledge graphs, and at the same time be able to evolve as the technologies progress.