

\chapter{Representation Impact on Knowledge Graph Refactoring from a Reification Perspective}
\label{chapter:evolution}

Knowledge graphs are not immutable, as they are subject to modifications along with the ontology or network of ontologies that are related to. This need for evolution may be triggered by changes in the domain of knowledge or on the graph consumption requirements in downstream tasks. 
An example is the need to add information about a triple, thus making statements about statements (i.e., statement reification).
%Reification is usually used to incorporate metadata and provenance to existing triples~\cite{frey2019evaluation}.
%Throughout the years, different representations have been proposed for reifying statements in RDF 
%, such as Named Graphs~\cite{carroll2005namedgraphs}, N-Ary Relationships~\cite{naryw3c2006} or RDF-star~\cite{hartig2017foundations}. 
For instance, Nanopublications~\parencite{groth2010nanopubs} adopt a Named Graph-based~\parencite{carroll2005namedgraphs} model to publish minimal scientific statements along with their provenance and associated context. Meanwhile, in ontology engineering, N-Ary Relationships~\parencite{naryw3c2006} are widely used as an ontology design pattern~\parencite{gangemi2013multi}. Moreover, the well-known knowledge base of Wikidata has implemented its own reification schema based on \textit{qualifiers}~\parencite{erxleben2014introducing}. 

It is not uncommon for an existing KG to strive for adopting one of these representations. Several studies throughout the years prove that these different representations impact the downstream KG consumption tasks; for query performance~\parencite{das2014tale,nguyen2014don,alocci2015property,hernandez2015reifying,frey2019evaluation,orlandi2021benchmarking} as well as for human consumption and machine learning-based tasks~\parencite{iglesias2023kgconsumption}. Some examples of existing KGs that have changed their representation include the preliminar transition of Wikidata to the current qualifier-based representation~\parencite{erxleben2014introducing}; the publication of the DisGeNet KG as Nanopublications~\parencite{queralt2016disgenet}, and the introduction of RDF-star in the European Union Agency for Railway  KG\footnote{\url{https://data-interop.era.europa.eu/}}.

When the situation comes that this kind of refactoring of the knowledge graph is needed, the following question arises: Is more convenient to construct the graph from the original input data sources again, or to re-construct it within the triplestore? This section analyses which of these approaches reports better performance and the parameters that influence the choice for a set of cases related to metadata representation approaches. 
To this end, we perform a comprehensive empirical study of this transformation in four representations (Standard Reification, N-Ary Relationships, Named Graphs and RDF-star) using (i) declarative mapping languages to construct each representation with KG construction engines and (ii) transformation per peers of representations with SPARQL \texttt{CONSTRUCT} queries in different triplestores. 
This chapter aims to unravel the aspects to consider when refactoring a knowledge graph from a reification perspective, to help in the decision on how to perform it and assess the role that mapping technologies can play in this kind of KG evolution. %We also raise awareness of recently overlooked aspects of \texttt{CONSTRUCT} queries, while contributing to research on the impact of RDF reification to enhance KG management.

\input{6_evolution/6-1_re-construction}

\section{Summary}

This chapter addresses the third objective of the thesis \textit{O3: To assess declarative knowledge graph construction technologies in terms of their benefits for supporting the creation and evolution of knowledge graphs}.
To this end, we study the re-construction of graphs with different technologies to implement different reification strategies. We evaluate, for a given KG where a refactoring of statement reification is needed, whether it is more efficient to construct the KG from the original data sources with the new representation, or to re-construct it within a triplestore. 
That is, to change the schema of a pre-existing knowledge graph without changing the data. To that end, we perform an empirical analysis using two KG construction engines and three triplestores, performing the re-construction with four data sizes in four different reification models. 
This comprises the main contribution associated with \textit{O3, C6: Analysis of the scenarios in which declarative KG construction technologies play a valuable role in the refactoring of knowledge graphs.}
%The results show that KG construction engines in general are more scalable, almost independently of the kind of reification. Triplestores perform best for small data sizes. However, their performance is highly dependent on the source representation (in the \texttt{WHERE} clause) and the target representation (in the \texttt{CONSTRUCT} clause). The time for producing RDF-star and Named Graphs is competitive with the KG construction approach. Yet, this approach becomes inefficient for producing N-Ary Relationships due to the need of introducing more joins in the \texttt{CONSTRUCT} clause; and when RDF-star is the source model, since most triplestores are not optimized for its new syntax. 

%We aim at providing a better understanding of re-constructing knowledge graphs from a reification perspective, contributing to the research on representations interoperability, and how mappings are a key element for scalable transitions. 

