\begin{itemize}
    \item The first contribution in \cref{chapter:mappings} was published in: \textbf{Iglesias-Molina, A.}, Cimmino, A., Ruckhaus, E., Chaves-Fraga, D., García-Castro, R., Corcho, O.: An Ontological Approach for Representing Declarative Mapping Languages, Semantic Web Journal 2022.
    
    \item The second contribution  in \cref{chapter:mappings} was preliminary published in: Delva, T., Arenas-Guerrero, J.,\textbf{ Iglesias-Molina, A.}, Corcho, O., Chaves-Fraga, D., and Dimou, A.: RML-star: A declarative mapping language for RDF-star generation, in Proceedings of the ISWC 2021 Posters, Demos and Industry Tracks (ISWC2021). The extended version of this contribution was published in: \textbf{Iglesias-Molina, A.}, Van Assche, D., Arenas-Guerrero, J., De Meester, B., Debruyne, C., Jozashoori, S., Maria, P., Michel, F., Chaves-Fraga, D. and Dimou, A., 2023. The RML Ontology: A Community-Driven Modular Redesign After a Decade of Experience in Mapping Heterogeneous Data to RDF, in Proceedings of the 22st International Semantic Web Conference (ISWC2023); and in: Arenas-Guerrero, J., \textbf{Iglesias-Molina, A.}, Chaves-Fraga, D., Garijo, D., Corcho, O. and Dimou, A., Declarative generation of RDF-star graphs from heterogeneous data, Semantic Web Journal, 2023 (the first two authors contributed equally to this publication). This contribution is the result of the collaboration with the Knowledge Graph Construction W3C Community Group. 

    \item The first contribution in \cref{chapter:creation} was initially published \ana{cambiar si se envia la ultima parte } in: \textbf{Iglesias-Molina, A.}, Chaves-Fraga, D., Priyatna, F. and Corcho, O., 2019, November. Towards the definition of a language-independent mapping template for knowledge graph creation, in Proceedings of the Third International Workshop on Capturing Scientific Knowledge co-located with the Eleventh International Conference on Knowledge Capture 2021 (K-CAP2021); and in: \textbf{Iglesias-Molina, A.}, Pozo-Gilo, L., Dona, D., Ruckhaus, E., Chaves-Fraga, D. and Corcho, O., Mapeathor: Simplifying the Specification of Declarative Rules for Knowledge Graph Construction, in Proceedings of the ISWC 2020 Demos and Industry Tracks (ISWC2020).

    \item The second contribution in \cref{chapter:creation} was published in: \textbf{Iglesias-Molina, A.}, Chaves-Fraga, D., Dasoulas, I. and Dimou, A., Human-Friendly RDF Graph Construction: Which One Do You Chose?, in Proceedings of the 23rd International Conference on Web Engineering 2023 (ICWE2023). This contribution is the result of the collaboration with KU Leuven.

    \item The first part of the contribution in \cref{chapter:evolution} was published 
    in: \textbf{Iglesias-Molina, A.}, Ahrabian, K., Ilievski, F., Pujara, J. and Corcho, O., Comparison of Knowledge Graph Representations for Consumer Scenarios, in Proceedings of the 22st International Semantic Web Conference (ISWC2023), as a result of a collaboration with the Information Sciences Institute (ISI-USC) thanks to a research stay in the institution.
    The second part of the contribution was published in: \textbf{Iglesias-Molina A.}, Toledo J., Corcho O. and Chaves-Fraga D., Re-Construction Impact on Metadata Representation Models, \ana{venue :)}, 2023.

\end{itemize}