\chapter{State of the Art}
\label{chapter:sota}

In this chapter, we discuss the state of the art of the topics of interest for this thesis, so as to provide an overview of recent developments in the field as well as to identify the open research problems and challenges in the area, some of which will be addressed in this thesis. The chapter starts describing the essential concepts of knowledge representation and annotation in the Web in \cref{sec:chp2_semweb}. Then, in \cref{sec:chp2_declarative_kgc} we focus on describing current techniques to construct knowledge graphs, particularly with declarative approaches. The challenges in the adoption of declarative KGC approaches lead us to present the user-friendly approaches developed to facilitate the creation of the mapping documents in \cref{sec:chp2_easy_kgc}. Next, \cref{sec:chp2_kg_lifecycle} introduces several knowledge graph life cycle proposals, analysing in which stages declarative mappings can play a beneficial role for its management. The chapter ends in \cref{sec:chp2_conclusions-sota} drawing the conclusions from each area of research and its main gaps, introducing how this thesis contributes to advance the state of the art. 


\input{2_stateoftheart/2-1_semweb}

\input{2_stateoftheart/2-2_declarative_kgc}

\input{2_stateoftheart/2-3_easy_kgc}

\input{2_stateoftheart/2-4_kg_lifecycle}

\section{Conclusions and Limitations of the State of the Art}
\label{sec:chp2_conclusions-sota}

This chapter presents the relevant work in the state of the art about declarative mapping languages for knowledge graph construction and evolution. We conclude in this section with the identification of the challenges in the field that motivate the contributions of this thesis, which are presented in the following chapters.

\begin{itemize}
    \item \textbf{Different mapping language expressiveness and processing characteristics of their associated engines.} Many different mapping languages for knowledge graph construction from heterogeneous data sources have been released over the years. These languages offer a differential set of features, providing a wide variety of possibilities for addressing real-world use cases. Deciding which language is the most suitable for the particularities of each use case and requirements becomes hard without a fine-grained comparison of the features that each language provides. 
    
    \item \textbf{Updating mapping languages according to new needs.} As KG technologies advance, new needs and requirements emerge in KGC. Despite the options that mapping languages currently provide, they have limitations and have not adopted yet some of the latest developments (e.g. the currently work in progress RDF 1.2 specification~\parencite{hartig2023rdf}). These limitations negatively impact on the adoption of declarative approaches for KGC, as they must be competitive with ad-hoc approaches (e.g. programming scripts) in terms of expressiveness and keep up with the current progress in the field.
    
    \item \textbf{Adoption barrier for writing declarative mappings.} Mappings usually follow a syntax that is not easy or intuitive to write, especially for non-experts. Several developments have been released, providing user interfaces and user-friendly serializations to improve the writing task for users. However, visual approaches have not been adopted in general, and serializations still pose a barrier for users with non-technical profiles. 

    %\item \ana{interoperability issue? not a contribution (in chp3) for now}
    
    \item \textbf{The role of declarative mappings in the knowledge graph life cycle.} Mapping languages were developed for constructing KGs from diverse data sources, relating them to the schema provided by an ontology. However, the analysis show that they can also be part of other tasks of the KG life cycle, such as data pre-processing and metadata annotation. Yet, the role that they can play in other parts of the life cycle has not been assessed. 
\end{itemize}
